`r glue::glue('Compare aggregated and non-aggregated reference: {params$library_id}')`

### 2025-07-03 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/01_reference-aggregation.Rmd
add UMAP with harmonized palette

---
title: "`r glue::glue('Compare aggregated and non-aggregated reference: {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  sample_id: "SCPCS000109"
  library_id: "SCPCL000126"
---

@@

validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
```

@@

@@

non_aggr <- readr::read_rds(
  file.path(
    results_dir,
    params$sample_id,
    "not-aggregated",
    glue::glue("{params$library_id}_singler-result.rds")
  )
) |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::mutate(ref_type = "not-aggregated")
```

@@

celltype_df <- celltype_df |>
  dplyr::left_join(singler_df, by = "barcodes") |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # set validation NA -> Unknown
  dplyr::mutate(
    validation_group_annotation =
      ifelse(is.na(validation_group_annotation), "Unknown", validation_group_annotation)
  )
```

@@

>>>>

---
title: "`r glue::glue('Compare aggregated and non-aggregated reference: {params$library_id}')`"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  sample_id: "SCPCS000105"
  library_id: "SCPCL000122"
---

@@

# broad consensus cell type groups
validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"

# umap colors
palette_url <- "https://raw.githubusercontent.com/AlexsLemonade/scpca-paper-figures/refs/heads/main/palettes/validation-group-palette.tsv"
```

@@

palette_df <- readr::read_tsv(palette_url) # umap plotting later

@@

non_aggr <- readr::read_rds(
  file.path(
    results_dir,
    params$sample_id,
    "not-aggregated",
    glue::glue("{params$library_id}_singler-result.rds")
  )
) |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::mutate(ref_type = "non-aggregated")
```

@@

celltype_df <- sce_df |>
  dplyr::left_join(singler_df, by = "barcodes") |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # set validation NA -> unknown
  dplyr::mutate(
    validation_group_annotation =
      ifelse(is.na(validation_group_annotation), "unknown", validation_group_annotation)
  )
```

@@

## UMAPs

Below, we show the UMAPs colored by cell types for each of aggregated, non-aggregated, and consensus (again, using the validation groupings for plot visibility).

Before plotting cell type labels were harmonized to the extent possible to that `NBAtlas` labels matched the corresponding consensus cell type labels (e.g., `Fibroblast` -> `fibroblast` or `Endothelial` -> `endothelial cell`).
Cell types which not literally just spelling variants were left alone.

The UMAP therefore has one overall legend for all possible cell types, so colors are directly comparable across panels.
The specific cell types that are part of either refernce are printed below.

```{r}
# Recode NBAtlas cell types where possible so they match with colors
celltype_recoded_df <- celltype_df |>
  dplyr::select(-consensus_annotation) |>
  # pivot longer for wrangling first
  tidyr::pivot_longer(
    c(validation_group_annotation, aggregated, not_aggregated),
    names_to = "annotation_type",
    values_to = "label"
  ) |>
  dplyr::mutate(
    label_recoded = dplyr::case_when(
      # nbatlas ~ consensus
      label %in% c("Myeloid", "Fibroblast") ~ stringr::str_to_lower(label),
      label == "Endothelial" ~ "endothelial cell",
      label == "Plasma" ~ "plasma cell",
      label == "NK cell" ~ "natural killer cell",
      is.na(label) ~ "unknown",
      .default = label
    )
  )

# Recode some existing colors we don't need
palette_df <- palette_df |>
  dplyr::mutate(
    label_recoded = dplyr::case_when(
      validation_group_annotation == "hematopoietic precursor cell" ~ "Neuroendocrine",
      validation_group_annotation == "muscle cell" ~ "RBCs",
      validation_group_annotation == "adipocyte" ~ "Schwann",
      validation_group_annotation == "pericyte" ~ "pDC",
      validation_group_annotation == "innate lymphoid cell" ~ "Stromal other",
      .default = validation_group_annotation
    )
  ) |>
  dplyr::filter(label_recoded %in% unique(celltype_recoded_df$label_recoded))

# Define palette
pal <- palette_df$color
names(pal) <- palette_df$label_recoded
```

```{r, fig.width = 12}
ggplot(celltype_recoded_df) +
  aes(x = UMAP1, y = UMAP2, color = label_recoded) +
  geom_point(alpha = 0.6, size = 0.5) +
  scale_color_manual(values = pal) +
  facet_wrap(vars(annotation_type)) +
  coord_equal() +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    legend.position = "bottom"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```
Cell types that are part of `NBAtlas` include:

```{r}
celltype_recoded_df |>
  dplyr::filter(annotation_type != "validation_group_annotation") |>
  dplyr::pull(label_recoded) |>
  unique()
```

Cell types that are part of consensus labels include:

```{r}
celltype_recoded_df |>
  dplyr::filter(annotation_type == "validation_group_annotation") |>
  dplyr::pull(label_recoded) |>
  unique()
```
